Random Matrix Theory Tools for the Analysis of Functional Magnetic Resonance Imaging Examinations

### 2022-10-31 paper/paper/SPIE/article.tex
WATCH OUT FOR CITEP

\section*{Approach}
We develop open-source software to efficiently compute RMT features from fMRI images, and examine
the cross-validated predictive potential of eigenvalue and RMT-based features ("eigenfeatures") used
with typical classificatory machine-learning algorithms. We systematically vary pre-processing extent,
normalization procedure, RMT unfolding procedure, and classifiers to examine the utility of these
eigenfeatures across various datasets.

@@

As our primary intention is to examine the potential utility of RMT-derived features across a wide
variety of data, and not to decisively put forward or refute any specific hypotheses relating RMT to
psychological processes, we examine a wide selection of fMRI datasets available on the
\href{https://openneuro.org/}{OpenNeuro} platform [36].  This allows for a broad assessment of the
potential of voxelwise RMT features.

However, since previous studies employing RMT for feature extraction have partially interpreted
their findings with respect to attentional processes
\citenum{wangRandomMatrixTheory2016,matharooSpontaneousBackpainAlters2020}, we have elected to use
datasets that involve attention in some way. This includes measures of attention or vigilance, or
comparisons between a condition involving more attention and focus (e.g. a task state) or conditions
which should involve considerably less attention or focus (e.g. resting-state).

@@

\begin{table}[h!]
\small
\centering
\begin{tabular}{ c c c c c c c }
\hline
\textbf{ID}    & \textbf{Dimensions}  & \textbf{Voxel size (mm)} & \textbf{TR} & \textbf{Volumes} & \textbf{n\_scans} \\
\hline
AGING     & 74 x 74 x 32   & 3.0 x 3.0 x 4.0 & 2.0  & 300 & 62  \\
BILING    & 100 x 100 x 72 & 1.8 x 1.8 x 1.8 & 0.88 & 823 & 90  \\
DEPRESS   & 112 x 112 x 25 & 2.0 x 2.0 x 5.0 & 2.5  & 100 & 72  \\
LEARN     & 64 x 64 x 36   & 3.0 x 3.0 x 3.0 & 2.0  & 195 & 432 \\
OSTEO     & 64 x 64 x 36   & 3.4 x 3.4 x 3.0 & 2.5  & 300 & 74  \\
PARK      & 80 x 80 x 43   & 3.0 x 3.0 x 3.0 & 2.4  & 149 & 552 \\
PSYCH     & 128 x 128 x 70 & 1.5 x 1.5 x 1.5 & 3.0  & 300 & 90  \\
\hline
\end{tabular}
\caption{
    ID = Identifier for paper. TR = Time of Repetition (seconds).
    Time = total duration (minutes) of each scan. Dimensions listed as M x N x P,
    indicate P horizontal / axial slices each with dimensions M x N.
}
\label{table:1}
\end{table}

\begin{table}[h!]
\small
\centering
\begin{tabular}{ l l c c c }
\hline
\textbf{ID}  & \textbf{Subgroup}  & \textbf{Subjects}  & \textbf{Task} & \textbf{Scans per Subject} \\
\hline
AGING   &  older             & 28      & resting-state         & 1 \\
        &  younger           & 34      & resting-state         & 1 \\
\hline
BILING  &  bilingual          & 59      & resting-state        & 1 \\
        &  monolingual        & 33      & resting-state        & 1 \\
\hline
DEPRESS &  depression         & 51      & resting-state        & 1 \\
        &  control            & 21      & resting-state        & 1 \\
\hline
LEARN   &  task              & 24      & learn image sketches  & 16 \\
        &  rest              & 24      & resting-state         & 2 \\
\hline
OSTEO   &  duloxetine        & 19      & taking duloxetine     & 1 \\
        &  pain              & 37      & placebo               & 1 \\
        &  nopain            & 20      & control               & 1 \\
\hline
PARK    &  Parkinson's       & 25      & ANT                   & 12 \\
        &  control           & 21      & ANT                   & 12 \\
\hline
PSYCH   &  vigilant           & 11     & resting-state        & 3 \\
        &  nonigilant         & 11     & resting-state        & 3 \\
        &  trait-attentive     & 11    & resting-state        & 3 \\
        &  trait-non-attentive & 11    & resting-state        & 3 \\
        &  task-attentive     & 11     & resting-state        & 3 \\
        &  task-nonattentive  & 11     & resting-state        & 3 \\
\hline
\end{tabular}
\caption{
    ID = Identifier for paper. FOV = Field of View. TR = Time of Repetition (seconds).
    Time = total duration (minutes) of each scan. Dimensions listed as M x N x P,
    indicate P slices with dimensions M x N.
}
\label{table:2}
\end{table}

@@

\subsection{Data, Materials, and Code Availability}
Availability of data, materials, and/or code used in the research results reported in the manuscript may be declared under the heading "Data, Materials, and Code Availability," following the Acknowledgments section. As relevant, provide specific access information or restrictions for data, materials, and computer code (i.e., links to repository access addresses with guidance on commercial or public access).

@@

For books\cite{Lamport94,Alred03,Goossens97} the listing includes the list of authors (initials plus last name), book title (in italics), page or chapter numbers, publisher, city, and year of publication.  Journal-article references \cite{Metropolis53,Harris06} include the author list, title of the article (in quotes), journal name (in italics, properly abbreviated), volume number (in bold), inclusive page numbers or citation identifier, and year.  A reference to a proceedings paper or a chapter in an edited book\cite{Gull89a} includes the author list, title of the article (in quotes), conference name (in italics), editors (if appropriate), volume title (in italics), volume number if applicable (in bold), inclusive page numbers, publisher, city, and year.  References to an article in the SPIE Proceedings may include the conference name, as shown in Ref.~\citenum{Hanson93c}.

The references are numbered in the order of their first citation. Citations to the references are made using superscripts, as demonstrated in the preceding paragraph. One may also directly refer to a reference within the text, for example, ``as shown in Ref.~\citenum{Metropolis53} ...''  Two or more references should be separated by a comma with no space between them. Multiple sequential references should be displayed with a dash between the first and last numbers \cite{Alred03,Perelman97,Lamport94,Goossens97,Metropolis53}.

>>>>

\section*{Approach}
We develop open-source software to efficiently compute RMT features from fMRI images, and examine
the cross-validated predictive potential of eigenvalue and RMT-based features (``eigenfeatures'') used
with typical classificatory machine-learning algorithms. We systematically vary pre-processing extent,
normalization procedure, RMT unfolding procedure, and classifiers to examine the utility of these
eigenfeatures across various datasets.

@@

As our primary intention is to examine the potential utility of RMT-derived features across a wide
variety of data, and not to decisively put forward or refute any specific hypotheses relating RMT to
psychological processes, we examine a wide selection of fMRI datasets available on the
\href{https://openneuro.org/}{OpenNeuro} platform (36).  This allows for a broad assessment of the
potential of voxelwise RMT features.

However, since previous studies employing RMT for feature extraction have partially interpreted
their findings with respect to attentional processes
\citenum{wangRandomMatrixTheory2016,matharooSpontaneousBackpainAlters2020}, we have also elected to use
datasets that involve attention in some way. This includes measures of attention or vigilance, or
comparisons between a condition involving more attention and focus (e.g. a task state) or conditions
which should involve less attention or focus (e.g. resting-state, pain medication vs. none).

\subsection{Dataset: PSYCH}
The PSYCH Dataset  is detailed in \citenum{gorgolewskiHighResolution7Tesla2015}. Subjects completed a
battery of psychological measurements, and various resting-state fMRI (rs-fMRI) whole-brain 7-Tesla
scans were obtained. For the purpose of keeping with the ``attention'' theme of the analyses in this
paper, three pairings of subgroups were constructed via z-score splits of the various sum-scores of
scale items involving concepts closely related to attention.

\emph{Trait Attention}.
The PANAS-X \citenum{watsonPANASXManualPositive1994} is a 9-point Likert-type self-report measure that
queries the extent to which subjects agree particular affective terms have applied to them over a
recent time period (e.g. a few weeks). An ad hoc (i.e. face-valid) scale was constructed from the
scale items: ``attentive'', ``alert'', ``interested'', ``concentrating'', and reverse-scored items
``sluggish'', ``tired'', ``sleepy'', ``drowsy''. Subjects were then placed in the high / low ``trait
attention'' category based on a median split of the average z-score across items.

\emph{Task Attention}.
The Conjunctive Continuous Performance Task (CCPT;
\citenum{shalevConjunctiveContinuousPerformance2011} is a behavioural measure of sustained attention
where subjects must react only if they see a red square among other different coloured figures. For
this paper, a sum score was constructed by subtracting the number of false positive and false
negative responses from the number of success responses, and subjects with a z-score greater / less
than zero on this scale were placed in high / low ``task attention'' category.

\emph{Vigilance}.
The Mini New York cognition questionnaire assesses various elements of the subject's cognitive
state, e.g. ``my thoughts involved other people'', ``I thought about something positive'' (translated
from German; \citenum{gorgolewskiCorrespondenceIndividualDifferences2014}. One question assesses
alertness (``I was fully awake'' / ``war ich vollkommen wach'') and was used to create high / low
``vigilance'' conditions in the same manner as the previous two splits.

\subsubsection{Dataset 2}
Dataset 2 (referred to in figures and tables with the label ``OSTEO'') is detailed in
\citenum{tetreaultBrainConnectivityPredicts2016}. Subjects included healthy controls (``nopain''
condition) and individuals with knee osteoarthritis (``pain'' condition) treated for 2 weeks with
either placebo (``pain'' condition) or duloxetine (``duloxetine'' condition). An additional grouping was
also created with both the placebo and duloxetine data (``allpain'' condition), allowing for six
subgroup comparisons. All subjects were imaged with whole-brain rs-fMRI. Scan parameters are
documented in Table \ref{table:1}. The use of rs-fMRI and RMT for examining pain makes this dataset
complementary to \citenum{matharooSpontaneousBackpainAlters2020}.

\subsubsection{Dataset 3}
Dataset 3  (referred to here with the label ``REFLECT'', due to the ``reflective” nature of the
cognitive tasks) is described in full detail by \citenum{dupreMultiechoFMRIReplication2016}. Both
multi-echo rs- and task-fMRI scans were conducted with scan parameters summarized in Table \ref{table:2}. Tasks
involved viewing various affective images (see \citenum{sprengPatternsBrainActivity2010} for details)
and performing tasks related to autobiographical memory, prospection, or theory-of-mind reasoning.
Continuing with the theme of broadly comparing high versus low attention throughout this paper, the
different tasks were simply treated generically as ``attending'', in contrast to the non-attentive
resting-state scans. This resulted in four task state scans and one resting-state scan per subject.

In order to combine the multi-echo data into a single, usable matrix, two methods were employed. In
the ``summed'' condition, echoes were combined via summation, a method which has value in providing an
image with overall increased contrast \citenum{kunduMultiechoFMRIReview2017}. In the ``interleave''
condition, echoes were interleaved at each voxel: that is, if we denote the voxel's \(i\)th signal value
from echo \(j\) as \(t_{j,i}\), then the interleaved voxel of the new image has the time-series \([t_{1,1},
t_{2,1}, t_{3,1}, \dots, t_{1,i}, t_{2,i}, t_{3,i}, \dots, t_{1,n}, t_{2,n}, t_{3,n}]\).

This creates a physically-meaningful 4D image that contains the full, original multi-echo fMRI
time-series data without any alterations. It should be noted that the purpose of these combination
methods is not to create a 4D volume that allows optimal detection of the purely task-linked BOLD
signal, but to ensure all scan information was used with minimal additional analytic assumptions.

\subsubsection{Dataset 4}
Dataset 4 (referred to in figures and tables with the label ``PARK'' or ``PARKINSONS'') is documented and
described in detail in \citenum{madhyasthaDynamicConnectivityRest2015}. Scan parameters are documented
in Table \ref{table:1}. Subjects with non-demented Parkinson's Disease (PD) and healthy controls performed a
number of repetitions of the Attention Network Task
(ANT; \citenum{fanActivationAttentionalNetworks2005} ). Subgroup details are available in Table \ref{table:1}. As the ANT
task demands high attention, it is not as straightforward to create ``high'' or ``low'' attention
groups. Instead, since attention presumably is high and varies little in this dataset, and subgroups
instead differ by virtue of the presence or absence of PD, any results here serve as a useful
comparison point for the results from studies that should clearly involve attentional differences.
This dataset also included extensively preprocessed scans, which were included in analyses to
examine the effects of preprocessing on the extracted RMT features.

\subsubsection{Dataset 5}
Dataset 5 (referred to in figures and tables with the label ``LEARN'') is documented and described in
detail in \citenum{schapiroHumanHippocampalReplay2018}. In this study, both rs-fMRI and task fMRI scans
were available for all subjects, allowing for within-subject comparisons. The complex task involved
multiple phases, wherein subjects were exposed to named, black-and-white outlined satellite images
related by family-resemblance of the components see figure 1 of \citenum{schapiroHumanHippocampalReplay2018}. After this learning task, subjects were tested on their
memory of the various learned features. Scans corresponding to both the initial learning and
subsequent testing phase were included in the high-attention ``task'' subgroup for the purposes of
the analysis of this paper. Resting state scans were included in the low-attention ``rest''
subgroup.

@@

\begin{table}[h!]
\caption{
    ID = Identifier for paper. TR = Time of Repetition (seconds).
    Time = total duration (minutes) of each scan. Dimensions listed as M x N x P,
    indicate P horizontal / axial slices each with dimensions M x N.
}
\label{table:1}
\small
\centering
\begin{tabular}{ c c c c c c c }
\hline
\textbf{ID}    & \textbf{Dimensions}  & \textbf{Voxel size (mm)} & \textbf{TR} & \textbf{Volumes} & \textbf{n\_scans} \\
\hline
AGING     & 74 x 74 x 32   & 3.0 x 3.0 x 4.0 & 2.0  & 300 & 62  \\
BILING    & 100 x 100 x 72 & 1.8 x 1.8 x 1.8 & 0.88 & 823 & 90  \\
DEPRESS   & 112 x 112 x 25 & 2.0 x 2.0 x 5.0 & 2.5  & 100 & 72  \\
LEARN     & 64 x 64 x 36   & 3.0 x 3.0 x 3.0 & 2.0  & 195 & 432 \\
OSTEO     & 64 x 64 x 36   & 3.4 x 3.4 x 3.0 & 2.5  & 300 & 74  \\
PARK      & 80 x 80 x 43   & 3.0 x 3.0 x 3.0 & 2.4  & 149 & 552 \\
PSYCH     & 128 x 128 x 70 & 1.5 x 1.5 x 1.5 & 3.0  & 300 & 90  \\
\hline
\end{tabular}

\end{table}

\begin{table}[h!]
\caption{
    ID = Identifier for paper. Subgroup = name of subgroup. Subjects = Number of subjects.
    Task = fMRI task. ANT = Attention Network Task \citenum{fanActivationAttentionalNetworks2005}
}
\label{table:2}
\small
\centering
\begin{tabular}{ l l c c c }
\hline
\textbf{ID}  & \textbf{Subgroup}  & \textbf{Subjects}  & \textbf{Task} & \textbf{Scans per Subject} \\
\hline
AGING   &  older             & 28      & resting-state         & 1 \\
        &  younger           & 34      & resting-state         & 1 \\
\hline
BILING  &  bilingual          & 59      & resting-state        & 1 \\
        &  monolingual        & 33      & resting-state        & 1 \\
\hline
DEPRESS &  depression         & 51      & resting-state        & 1 \\
        &  control            & 21      & resting-state        & 1 \\
\hline
LEARN   &  task              & 24      & learn image sketches  & 16 \\
        &  rest              & 24      & resting-state         & 2 \\
\hline
OSTEO   &  duloxetine        & 19      & taking duloxetine     & 1 \\
        &  pain              & 37      & placebo               & 1 \\
        &  nopain            & 20      & control               & 1 \\
\hline
PARK    &  Parkinson's       & 25      & ANT                   & 12 \\
        &  control           & 21      & ANT                   & 12 \\
\hline
PSYCH   &  vigilant           & 11     & resting-state        & 3 \\
        &  nonigilant         & 11     & resting-state        & 3 \\
        &  trait-attentive     & 11    & resting-state        & 3 \\
        &  trait-non-attentive & 11    & resting-state        & 3 \\
        &  task-attentive     & 11     & resting-state        & 3 \\
        &  task-nonattentive  & 11     & resting-state        & 3 \\
\hline
\end{tabular}
\end{table}

@@

\subsection{Data, Materials, and Code Availability}
Availability of data, materials, and/or code used in the research results reported in the manuscript may be declared under the heading ``Data, Materials, and Code Availability,'' following the Acknowledgments section. As relevant, provide specific access information or restrictions for data, materials, and computer code (i.e., links to repository access addresses with guidance on commercial or public access).

@@

For books\citenum{Lamport94,Alred03,Goossens97} the listing includes the list of authors (initials plus last name), book title (in italics), page or chapter numbers, publisher, city, and year of publication.  Journal-article references \citenum{Metropolis53,Harris06} include the author list, title of the article (in quotes), journal name (in italics, properly abbreviated), volume number (in bold), inclusive page numbers or citation identifier, and year.  A reference to a proceedings paper or a chapter in an edited book\citenum{Gull89a} includes the author list, title of the article (in quotes), conference name (in italics), editors (if appropriate), volume title (in italics), volume number if applicable (in bold), inclusive page numbers, publisher, city, and year.  References to an article in the SPIE Proceedings may include the conference name, as shown in Ref.~\citenum{Hanson93c}.

The references are numbered in the order of their first citation. Citations to the references are made using superscripts, as demonstrated in the preceding paragraph. One may also directly refer to a reference within the text, for example, ``as shown in Ref.~\citenum{Metropolis53} ...''  Two or more references should be separated by a comma with no space between them. Multiple sequential references should be displayed with a dash between the first and last numbers \citenum{Alred03,Perelman97,Lamport94,Goossens97,Metropolis53}.
